Multi-Domain Adaptation in Neural Machine Translation with Dynamic Sampling Strategies

### 2021-05-15 2021/emnlp/auto_cl_mdmt.tex
Now the right length

% === The part on multidomain training has been moved to the end.

@@

\begin{table*}[htbp]
  \centering \small
  \begin{tabular}{|l|*{10}{r|}} \cline{1-2} \cline{4-11}
    domain \hfill $d=$ & \multicolumn{1}{c|}{\domain{news}}& \hfill &domain \hfill $d=$ & \multicolumn{1}{c|}{\domain{ med}} & \multicolumn{1}{c|}{\domain{ law}} & \multicolumn{1}{c|}{\domain{bank}} & \multicolumn{1}{c|}{\domain{talk}} & \multicolumn{1}{c|}{\domain{ it }} & \multicolumn{1}{c|}{\domain{ rel}} & \multicolumn{1}{c|}{mean}  \\ 
\cline{1-2} \cline{4-11}
    \multicolumn{2}{|c|}{\sl Unseen domain} & &\multicolumn{8}{c|}{\sl Training with 30 clusters} \\ 
\cline{1-2} \cline{4-11}
    \system{Mixed-0}      & 25.7 & &\system{DDS($\vlambda^*, \vlambda_d$)}&38.3&60.1&50.3&35.8&49.1&90.1&53.9\\
    \system{Mixed-0.25} & 25.8 & &\system{MDAC($\vlambda^*, \vlambda_d$)}&39.2$^*$&61.6$^*$&52.0$^*$&38.2$^*$&49.1&89.7&55.0$^*$\\ \cline{4-11}
    \system{Mixed-0.5}   &26.5\\
    \system{Mixed-0.75} &26.8\\
    \system{Mixed-1} &26.9 \\
    \cline{1-2}
     \system{DDS($\vlambda_0, \vlambda_{news}$)} &26.3 \\
     \system{MDAC($\vlambda_0, \vlambda_{news}$)} &26.3 \\
     \cline{1-2}
  \end{tabular}
  \caption{Few shot adaptation (left) and unsupervised adaptation (right). For a given line, each column correspond to one distinct system. ($^*$)MDAC is significantly better than DDS.}
  \label{tab:unsupervised-da}
\end{table*}

\subsection{Unseen domain}\label{ssec:uda}
The left part of Table~\ref{tab:unsupervised-da} displays the performance on the unseen domain \domain{news} for NMT systems trained with the mixtures $\lambda^l \in \big[ \lambda_0, \lambda_{0.25}, \lambda_{0.5}, \lambda_{0.75}, \lambda_{1.0}\big]$ and wiht dynamic data selection (MDAC and DDS). These systems have insignificant differences in BLEU, showing that dynamic mixture does not improve the robustness of the NMT system against unseen domains. However, the fact that the performance of MDAC and DDS is close to the best performance is a sign that they can also apply in such challengin situation. % the robustness of the NMT model.

\subsection{Automatic clustering}\label{ssec:clda}
The right part of Table~\ref{tab:unsupervised-da} reports the performance of NMT systems adapted to each domain. In comparison to section~\ref{ssec:da}, the training data is distributed in 30~clusters instead of the 6 original domains. Spliting the train data in small groups provides the learner with extra degrees of freedom when selecting the best distribution. However, as these clusters are built automatically, they are more noisy in nature. According to the results in Table~\ref{tab:unsupervised-da}, this scenario is hard both for DDS and MDAC which perform much worse than for the supervised DA setting. This again signals the importance of initialization: analyzing the clustering, we find that the data for \domain{rel} mostly correspond to one single cluster. Using a uniform initialization, this cluster starts with a very small weight, and never succeeds to yield the (good) performance observed in the DA settting.

@@

Domain adaptation is an ancient and vexing problem that has been studied from many angles both for SMT and NMT. A survey of supervised and unsupervised DA for NMT is in \citet{Chu18asurvey}, where they distinguish between data-centric and model-centric DA, a view also adopted in the more recent review of \citet{Saunders21domain}. Our approach to DA in this paper clearly falls under the former category. We refer interested readers to these references and do not discuss DA any further.

Multi-domain adaptation (MDA) aims to develop systems that simultaneously bode well for several domains. Like for DA, techniques for supervised MDA typically combine one or several ingredients: (a) the specialization of data representations \citep{Kobus17domaincontrol} or of sub-networks \citep{Pham19generic} to differentiate the processing of each domain; (b) the use of adversarial techniques to neutralize differences between domains \citep{Britz17mixing,Zeng18multidomain}; (c) the use of automatic domain identification e.g.\ \citet{Jiang19multidomain}. Unsupervised MDA is studied in \citet{Farajian17multidomain}, as an instance of unsupervised DA.

Most approaches to adaptive / dynamic data selection take inspiration from \citet{Bengio09curriculum}, where the notion of curriculum learning is initially introduced. CL relies on the notion of ``easiness'' of a sample to schedule the presentation of the training data so that the easiest examples presented first and the hardest, last. Various ways to automate CL using the framework of multi-armed bandits are notably explored in \citep{Graves17automated}, which has been an inspiration for our implementation. While the initial aim was primarily to improve and speed up training, CL has also proven useful for domain adaptive / multi-domain / multi-lingual MT, based on alternative definitions of ``easiness''.\fyDone{Ajouter Grave}

For instance, \citet{Zhang19curriculum} study supervised domain adaptation and propose a curriculum approach which progressively extends the data used in training: in the early stages only in-data is used, while shards containing less relevant\footnote{Domain distance is computed with Lewis-Moore scores (based on the cross-entropy of in-domain LM).} data are introduced in later stages. This is somehow opposite to the recommendations of \citet{Vanderwees17dynamic}, whose \emph{gradual fine-tuning} progressively focuses on the in-domain data.\fyDone{These have not been compared? and also to what we do ?}

\citet{Kumar19reinforcement} use reinforcement techniques (deep Q-learning) to learn the curriculum strategy: in this study, complexity corresponds to difficulty levels and which are binned based on contrastive data selection. The reward is based on the increase of the development set loss that results from the actual data selection strategy.\fyDone{Alert: what do we do during warm-up ?} The same technique is recently applied to the training of a multilingual NMT system \citep{Kumar21learning}. \citet{Zhou20uncertainty} propose another curriculum-based approach which instead relies on \emph{instance uncertainty} as a measure of their difficulty and presents the data sample starting with the easiest (more predictable) first. Another contribution of this paper is an alternative criterium for stopping the training. More related to our problems, \citet{Wang20learning-multi} adapt curriculum learning for multi-domain adaptation, where an optimal instance weighting scheme is found using Bayesian optimization (derivative-free) techniques. Each step consists of (a) weighting the instances based on relevance features, (b) fine-tuning a pretrained model using the (weighted) training set, and is applied iteratively to generate a sequence of models. The one that maximizes the development set performance is finally retained.

>>>>

@@

\begin{table*}[htbp]
  \centering \small
  \begin{tabular}{|l|*{10}{r|}} \cline{1-2} \cline{4-11}
    domain \hfill $d=$ & \multicolumn{1}{c|}{\domain{news}}& \hfill &domain \hfill $d=$ & \multicolumn{1}{c|}{\domain{ med}} & \multicolumn{1}{c|}{\domain{ law}} & \multicolumn{1}{c|}{\domain{bank}} & \multicolumn{1}{c|}{\domain{talk}} & \multicolumn{1}{c|}{\domain{ it }} & \multicolumn{1}{c|}{\domain{ rel}} & \multicolumn{1}{c|}{mean}  \\ 
\cline{1-2} \cline{4-11}
    \multicolumn{2}{|c|}{\sl Unseen domain} & &\multicolumn{8}{c|}{\sl Training with 30 clusters} \\ 
\cline{1-2} \cline{4-11}
    \system{Mixed-0}      & 25.7 & &\system{DDS($\vlambda^*, \vlambda_d$)}&38.3&60.1&50.3&35.8&49.1&90.1&53.9\\
    \system{Mixed-0.25} & 25.8 & &\system{MDAC($\vlambda^*, \vlambda_d$)}&39.2$^*$&61.6$^*$&52.0$^*$&38.2$^*$&49.1&89.7&55.0$^*$\\ \cline{4-11}
    \system{Mixed-0.5}   &26.5\\
    \system{Mixed-0.75} &26.8\\
    \system{Mixed-1} &26.9 \\
    \cline{1-2}
     \system{DDS($\vlambda_0, \vlambda_{news}$)} &26.3 \\
     \system{MDAC($\vlambda_0, \vlambda_{news}$)} &26.3 \\
     \cline{1-2}
  \end{tabular}
  \caption{Few shot adaptation (left) and unsupervised adaptation (right). For a given line, each column correspond to one distinct system. ($^*$) MDAC is significantly better than DDS.}
  \label{tab:unsupervised-da}
\end{table*}

\subsection{Unseen domain}\label{ssec:uda}
The left part of Table~\ref{tab:unsupervised-da} displays the performance on the unseen domain \domain{news} for NMT systems trained with the mixtures $\lambda^l \in \big[ \lambda_0, \lambda_{0.25}, \lambda_{0.5}, \lambda_{0.75}, \lambda_{1.0}\big]$ and wiht dynamic data selection (MDAC and DDS). These systems have insignificant differences in BLEU, showing that dynamic mixture does not improve the robustness of the NMT system against unseen domains. However, the fact that the performance of MDAC and DDS is close to the best performance is a sign that they can also apply in such challenging situation. % the robustness of the NMT model.

\subsection{Automatic clustering}\label{ssec:clda}
The right part of Table~\ref{tab:unsupervised-da} reports the performance of NMT systems adapted to each domain. In comparison to Section~\ref{ssec:da}, the training data is distributed in 30 automatic clusters instead of the 6 original domains. Spliting the train data in small groups provides the learner with extra degrees of freedom when selecting the best distribution. However, as these clusters are built automatically, they are more noisy in nature. According to the results in Table~\ref{tab:unsupervised-da}, this scenario is hard both for DDS and MDAC which perform much worse than for the supervised DA setting. This again signals the importance of initialization: analyzing the clustering, we find that the data for \domain{rel} mostly correspond to one single cluster. Using a uniform initialization, this cluster starts with a very small weight, and never succeeds to yield the (good) performance observed in the DA settting.

@@

Domain adaptation is an old problem that has been studied from many angles both for SMT and NMT. A survey of supervised and unsupervised DA for NMT is in \citet{Chu18asurvey}, where the authors distinguish between data-centric and model-centric DA, a view also adopted in the more recent review of \citet{Saunders21domain}. Our approach to DA in this paper falls under the former category. We refer interested readers to these references and do not discuss DA any further.

Multi-domain adaptation (MDA) aims to develop systems that simultaneously bode well for several domains. Like for DA, techniques for supervised MDA combine one or several ingredients: (a) the specialization of data representations \citep{Kobus17domaincontrol} or of sub-networks \citep{Pham19generic} to differentiate the processing of each domain; (b) the use of adversarial techniques to neutralize differences between domains \citep{Britz17mixing,Zeng18multidomain}; (c) the use of automatic domain identification e.g.\ \citet{Jiang19multidomain}. Unsupervised MDA is studied in \citet{Farajian17multidomain}, as an instance of unsupervised DA.

Most approaches to adaptive / dynamic data selection take inspiration from \citet{Bengio09curriculum}, where the notion of curriculum learning is introduced. CL relies on the notion of ``easiness'' of a sample to schedule the presentation of training data so that the easiest examples are presented first and the hardest, last. Various ways to automate CL using the framework of multi-armed bandits are explored in \citep{Graves17automated}, which has been an inspiration for our implementation. While the initial aim was primarily to improve and speed up training, CL has also proven useful for domain adaptive / multi-domain / multi-lingual MT, based on alternative definitions of ``easiness''.\fyDone{Ajouter Grave}
For instance, \citet{Zhang19curriculum} study supervised DA and propose a curriculum approach which progressively augments the training data: in the early stages, only in-data is used, while shards containing less relevant\footnote{Domain distance is computed with Lewis-Moore scores (based on the cross-entropy of in-domain LM).} data are introduced in later stages. This is somehow opposite to the recommendations of \citet{Vanderwees17dynamic}, whose \emph{gradual fine-tuning} progressively focuses on the in-domain data.\fyDone{These have not been compared? and also to what we do ?}

\citet{Kumar19reinforcement} use reinforcement techniques (deep Q-learning) to learn the curriculum strategy: in this work, complexity corresponds to difficulty levels which are binned using contrastive data selection. The reward is based on the increase of the development set loss that results from the actual data selection strategy.\fyDone{Alert: what do we do during warm-up ?} The same technique is recently applied to multilingual NMT in \citep{Kumar21learning}. \citet{Zhou20uncertainty} propose another curriculum-based approach which instead relies on \emph{instance uncertainty} as a measure of their difficulty and presents the data sample starting with the easiest (more predictable) first. Another contribution of this paper is an alternative criterium for stopping the training. More related to our problems, \citet{Wang20learning-multi} adapt CL for multi-domain adaptation, where an optimal instance weighting scheme is found using Bayesian optimization techniques. Each step consists of (a) weighting instances based on relevance features, (b) fine-tuning a pretrained model using the weighted training set, and is applied iteratively to train a sequence of models. The one that maximizes the development set performance is finally retained.
